A new method to estimate dark matter halo concentrations

### 2015-01-26 paper/article.tex
fixed some typos

We present a new method to estimate the concentration of dark matter
halos in N-body simulations. Our method is based on a fit to the
integrated mass density as a function of halo radius. 
The advantage of this method is that the full particle information
without any binning. 
We test our method both on mock and N-body halos to compare it against
two popular methods to find concentrations: radial particle binning to
estimate the density and the maximum radial velocity. 
Tests on the mock halos show that the accuracy of our method to
recover input concentrations can achieve $0.1\%$ whil for the maximum
radial velocity and density methods it is on the order of $0.5\%$ and
$40\%$ respectively. 
We also measure the mass-concentration relationship on the N-body data with
the three methods. With respect to the other two approaches our method
produces a flatter relationship, while its normalization falls below
the maximum velocity method and above the density method.

@@

One of the most striking results of these simulations is that dark
matter clumps on galactic scales follow a universal density profile
which in a first approximation is spherical symmetric and only
dependent on the radial coordinate. These profiles seem to be
universal; independent of the cosmological parameters and self-similar
for different spatial scapes after an addecuate re-scaling is
appplied.

One the most commong parameterization of this density is known as the
Navarro-Frenk-White (NFW) profile \citep{NFW}. This profile is a
double power law in radius, where the transition between the two
happens at the so-called scale radius $r_s$. The ratio between the
scale radius and the virial radius of the halo $R_v$ is known as the
concentration $c=R_V/r_s$, a quantity that has been found to be
correlated with the halo mass.

The relationship between halo mass and concentration is in principle
accesible to observations and provides a potential test of LCDM on
galactic scales. With this promise a great deal of effort has been
invested in calibrating this relationship with simulations but also
finding the best possible way to constraint it with observations.

From the computational point of view there are two main methods to
estimate the concetration parameter of a dark matter halo in a N-body
simulation. The first method takes the particles comprosing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of radius. A second method uses an analytical property of the
NFW that related the maximum of the ratio of the circular velocity to
the virial velocity. In this method at each particle radius this ratio
is estimated to find the maximum value for this ratio; this value is
used to find the roots of an equation which represent the
concentration parameter.

@@

Our method consists in building the cumulative mass profile from the
particle data in the N-body simulation to find the best possible
concentration value using a Markov Chain Monte Carlo methodology by
comparing the data against the analytical expectation.

This paper is structured as follows. In Section \ref{sec:basics} we
review the basic properties of the NFW density profile and define the
basic notation for the rest of the paper. Next in Section
\ref{sec:method} we present our new method to estimate the halo
concentration, payin special attention to the bayesian framework to
find the most probable value and its uncertaintt. In Section
\ref{sec:halo_sample} we demonstrate the power of our method with two
different halo samples; the first one generated under controlled
conditions and the second taken from a cosmological N-body
simulation. Next in \ref{sec:discussion} we discuss these results by
comparing them against other methods to estimate the concentration and
comment on the consequences for the concentration-mass
relationship. We present our conclusions in Section \ref{sec:conclusions}.

@@

\begin{equation}
v(<x)=\sqrt{\frac{1}{A}\left[\frac{\ln\left(1+xc\right)}{x}-\frac{c}{xc+1}\right]},
\end{equation}
%
this normalized profile always shows a maximum provided that the
concentration is larger than a values of $c>???$.
It is possible to show that for the NFW profile the maximum is
provided by

@@

The first method tries to directly estimate the density profile.
It takes all the particles in the halo and bins them in the logarithm
of the radial coordinate from the halo center.
Then, it estimates the density in each logarithmic bin counting the
particles and divinding by the corresponding shell volume.
At this point is is possible to make a direct fit to the density as a
function of the radial coordinate.
This method has been most recently used by \cite{Ludlow2014} to study
the mass-concentration-redshift relation of dark matter halos using
the Millennium Simulation Series.

A second method uses the circular velocity profile.
As it was shown in Fig. (\ref{fig:velocity}) the circular velocity
shows a maximum for all profiles with concentration values larger than
$c>2$.
The method finds the value of $x$ for which the normalized circular
velocity $v(<x)$ shows a maximum.
Using this value it solves numerically for the corresponding value of
the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2014} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

The value of $r_i$ is the radius of the $i$-th particle of the mock
halo.
Then we generate random polar and azimuthal angles $\theta$ and $\phi$
for each particle to ensure spherical symmetry.
Finally these three spherical coordinates are transformed into cartesian coordinates
$(r,\theta,\phi) \rightarrow (x,y,z)$.

We generate in total $400$ mock halos splitted into four different
groups of $100$ halos each.
The four groups differ in the total number of particles for their halos:
$20$, $200$, $2000$ and $20000$.
Inside each group the halos have random concentration values in
the range $1<c<20$ with a flat distribution.
For all these halos with find the concentration values using the
density, velocity and mass methods described in the previous
section. We quantify the difference between the expected $c_{in}$
and obtained $c_{out}$ values by

@@

Figure \ref{fig:results_mocks} shows the integrated distribution for
$D$ for the fits using our method, splitted into four different groups
according the particle number.
From this Figure the first immediate
conclusion is that increasing the number of particles increases the
chances to recover the input values.

We believe that the main effect that contributes to this trend
is that the particle that our algorithm finds to be the halo center
(where the potential is minimum) gets closer to the original
geometrical center (where no particle sits by construction) used to
generate the halo.  Poissonian noise makes this center fluctuate,
changing the numerical radial profile from the analytical input.

For particle number of $20$ the offset betwen the input and output
concentration can be as large as $20\%$, with a slight bias around
$-0.05\%$, i.e. the output concentration is biased towards lower
values than the input.
For particle numbers of $2000$ most of the offsets fall below $5\%$,
with a clear peak around $0\%$ indicating that any appreciable bias is
absent.

@@

Using this dataset we also compare our method against the other two
methods described earlier: using shells to estimate the density as a
function of radius and the maximum circular velocity method.
In the first case we use the same Metropolis-Hastings algorithm we
have in our methot to fit the density profile.
In the second method we simply follow the procedure described in
Section \ref{sec:method}

@@

Figure \ref{fig:error} shows the behaviour of $\avg{|D|}$ as a function of
halo particle number for the three different methods to estimate the
concentration.
At fixed particle numbers our method always shows the lowest
$\avg{|D|}$ values compared to the other two methods.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
going down to $0.1\%$ for halos with $20000$ particles.
The decrease of $\avg{|D|}$ with increasing particle number $N$ goes
approximately as $\avg{|D|}\propto N^{-1/2}$, which reinforces the hint that
the accuracy of the method is related to a decrease of Poissonian
noise.

The second best accuracy is achieved by the method based on the
maximum of the circular velocity.
It also shows a similar behaviour $\avg{|D|}\propto N^{-1/2}$.
Its accuracy is $2-5$ times less than in our method, on the order of
$20\%$ for $20$ particle halos and $0.5\%$ for $20000$ particle halos.
The method based on the direct density fit shows the lowest accuracy
of all methods.
Its average values $\avg{|D|}$ are always between $30\%$ adn $50\%$
regardless of the particle number. {\bf Necesitamos entender el
  comportamiento de esto.}

@@

We build a sample of all halos located in a cubic sub-volume of $100$
\hMpc on a side centered on the most massive halo in the simulation at
$z=0$ which corresponds to the \texttt{miniMDR1} tables in the
database.
From this sample we select all the halos at $z=0$ detected with a
Friends-of-Friends (FoF) algorithm with masses in the interval
$10^{11}\leq M_{\rm FoF}/\hMsun \leq 10^{15}$.
The FoF algorithm used with a linking length of $0.17$ times the average
interparticle distance. This choice translates into an overdensity
$\Delta_h\sim 400-700$ dependent on the halo concentration
\citep{More2011}.
Finally, for each FoF halo we select from all the particles that
belong to it.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.90\textwidth]{concentration.pdf}
\end{center}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. The central lines
  correspond the median and the shadowed region indicates the
  quartiles.
    \label{fig:concentrations}}
\end{figure*}

>>>>

We present a new method to estimate the concentration of dark matter
halos in N-body simulations. Our method is based on a fit to the
integrated mass as a function of halo radius. 
The main advantage of this method is that it uses the full particle information
without any binning. 
We test our method both on mock and N-body halos to compare it against
two popular methods to find concentrations: maximum radial velocity
measurements and radial particle binning to estimate the density.
Tests on the mock halos show that the accuracy of our method to
recover input concentrations can achieve $0.1\%$ while for the maximum
radial velocity and density methods it is on the order of $0.5\%$ and
$40\%$ respectively. 
We also measure the mass-concentration relationship on the N-body
data. With respect to the other two approaches our method produces a
flatter relationship, while its normalization falls below the maximum
velocity method and above the density method.

@@

One of the most striking results of these simulations is that dark
matter clumps on galactic scales follow a universal density profile
which in a first approximation is spherical symmetric and only
dependent on the radial coordinate. These profiles seem to be
universal; independent of the cosmological parameters and self-similar
for different spatial scales after an adequate re-scaling is
app plied.

One the most common parameterization of this density is known as the
Navarro-Frenk-White (NFW) profile \citep{NFW}. This profile is a
double power law in radius, where the transition between the two
happens at the so-called scale radius $r_s$. The ratio between the
scale radius and the virial radius of the halo $R_v$ is known as the
concentration $c=R_V/r_s$, a quantity that has been found to be
correlated with the halo mass.

The relationship between halo mass and concentration is in principle
accessible to observations and provides a potential test of LCDM on
galactic scales. With this promise a great deal of effort has been
invested in calibrating this relationship with simulations but also
finding the best possible way to constraint it with observations.

From the computational point of view there are two main methods to
estimate the concentration parameter of a dark matter halo in a N-body
simulation. The first method takes the particles composing the halos
and bins them in logarithmic radii to estimate the density in each
bin, then it proceeds with a fit of this density estimation as a
function of radius. A second method uses an analytic property of the
NFW that related the maximum of the ratio of the circular velocity to
the virial velocity. In this method at each particle radius this ratio
is estimated to find the maximum value for this ratio; this value is
used to find the roots of an equation which represent the
concentration parameter.

@@

Our method consists in building the cumulative mass profile from the
particle data in the N-body simulation to find the best possible
concentration value using a Markov Chain Monte Carlo methodology by
comparing the data against the analytic expectation.

This paper is structured as follows. In Section \ref{sec:basics} we
review the basic properties of the NFW density profile and define the
basic notation for the rest of the paper. Next in Section
\ref{sec:method} we present our new method to estimate the halo
concentration, paying special attention to the Bayesian framework to
find the most probable value and its uncertainty. In Section
\ref{sec:results} we demonstrate the power of our method with two
different halo samples; the first one generated under controlled
conditions and the second taken from a cosmological N-body
simulation to comment on the consequences for the concentration-mass
relationship. We present our conclusions in Section
\ref{sec:conclusions}.

@@

\begin{equation}
v(<x)=\sqrt{\frac{1}{A}\left[\frac{\ln\left(1+xc\right)}{x}-\frac{c}{xc+1}\right]},
\end{equation}
%
this normalized profile always shows a maximum provided that the
concentration is larger than $c>2$.
It is possible to show that for the NFW profile the maximum is
provided by

@@

The first method tries to directly estimate the density profile.
It takes all the particles in the halo and bins them in the logarithm
of the radial coordinate from the halo center.
Then, it estimates the density in each logarithmic bin counting the
particles and dividing by the corresponding shell volume.
At this point is is possible to make a direct fit to the density as a
function of the radial coordinate.
This method has been most recently used by \cite{Ludlow2014} to study
the mass-concentration-redshift relation of dark matter halos using
the Millennium Simulation Series.

A second method uses the circular velocity profile.
As it was shown in the right panel of Figure (\ref{fig:profiles}) the
circular velocity shows a maximum for all profiles with concentration
values larger than $c>2$.
The method finds the value of $x$ for which the normalized circular
velocity $v(<x)$ shows a maximum.
Using this value it solves numerically for the corresponding value of
the concentration using Eq. (\ref{eq:max_v}).
This method has been most recently used by \cite{Klypin2014} to study
the mass-concentration-redshift relation using the Multidark
Simulation Suite.

@@

The value of $r_i$ is the radius of the $i$-th particle of the mock
halo.
Then we generate random polar and azimuthal angles $\theta$ and $\phi$
for each particle to ensure spherical symmetry.
Finally these three spherical coordinates are transformed into Cartesian coordinates
$(r,\theta,\phi) \rightarrow (x,y,z)$.

We generate in total $400$ mock halos split into four different
groups of $100$ halos each.
The four groups differ in the total number of particles for their halos:
$20$, $200$, $2000$ and $20000$.
Inside each group the halos have random concentration values in
the range $1<c<20$ with a flat distribution.
For all these halos with find the concentration values using the
density, velocity and mass methods described in the previous
section. We quantify the difference between the expected $c_{in}$
and obtained $c_{out}$ values by

@@

Figure \ref{fig:results_mocks} shows the integrated distribution for
$D$ for the fits using our method, split into four different groups
according the particle number.
From this Figure the first immediate
conclusion is that increasing the number of particles increases the
chances to recover the input values.

We believe that the main effect that contributes to this trend
is that the particle that our algorithm finds to be the halo center
(where the potential is minimum) gets closer to the original
geometrical center (where no particle sits by construction) used to
generate the halo.  Poisson noise makes this center fluctuate,
changing the numerical radial profile from the analytic input.

For particle number of $20$ the offset between the input and output
concentration can be as large as $20\%$, with a slight bias around
$-0.05\%$, i.e. the output concentration is biased towards lower
values than the input.
For particle numbers of $2000$ most of the offsets fall below $5\%$,
with a clear peak around $0\%$ indicating that any appreciable bias is
absent.

@@

Using this data-set we also compare our method against the other two
methods described earlier: using shells to estimate the density as a
function of radius and the maximum circular velocity method.
In the first case we use the same Metropolis-Hastings algorithm we
have in our method to fit the density profile.
In the second method we simply follow the procedure described in
Section \ref{sec:method}

@@

Figure \ref{fig:error} shows the behaviour of $\avg{|D|}$ as a function of
halo particle number for the three different methods to estimate the
concentration.
At fixed particle numbers our method always shows the lowest
$\avg{|D|}$ values compared to the other two methods.
Its accuracy is on the order of $10\%$ for $20$ particles in the halo,
going down to $0.1\%$ for halos with $20000$ particles.
The decrease of $\avg{|D|}$ with increasing particle number $N$ goes
approximately as $\avg{|D|}\propto N^{-1/2}$, which reinforces the hint that
the accuracy of the method is related to a decrease of Poisson
noise.

The second best accuracy is achieved by the method based on the
maximum of the circular velocity.
It also shows a similar behaviour $\avg{|D|}\propto N^{-1/2}$.
Its accuracy is $2-5$ times less than in our method, on the order of
$20\%$ for $20$ particle halos and $0.5\%$ for $20000$ particle halos.
The method based on the direct density fit shows the lowest accuracy
of all methods.
Its average values $\avg{|D|}$ are always between $30\%$ and $50\%$
regardless of the particle number. {\bf Necesitamos entender el
  comportamiento de esto.}

@@

We build a sample of all halos located in a cubic sub-volume of $100$
\hMpc on a side centered on the most massive halo in the simulation at
$z=0$ which corresponds to the \texttt{miniMDR1} tables in the
database.
From this sample we select all the halos at $z=0$ detected with a
Friends-of-Friends (FoF) algorithm with masses in the interval
$10^{11}\leq M_{\rm FoF}/\hMsun \leq 10^{15}$.
The FoF algorithm used with a linking length of $0.17$ times the average
inter-particle distance. This choice translates into an overdensity
$\Delta_h\sim 400-700$ dependent on the halo concentration
\citep{More2011}.
Finally, for each FoF halo we select from all the particles that
belong to it.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.90\textwidth]{concentration.pdf}
\end{center}
\caption{Mass-concentration relationship for the three different
  methods used on the same cosmological N-body data. The central lines
  correspond the median and the shadowed region indicates the
  quartiles.
    \label{fig:concentration}}
\end{figure*}
